
\argomento{Diffusion Processes and the Need for Itô Calculus}

\textbf{Definition (Diffusion Process).}  
A \textbf{diffusion process} \( (X_t)_{t \geq 0} \) is a continuous-time stochastic process that satisfies a \textbf{stochastic differential equation} (SDE) of the form:
\[
dX_t = \mu(t, X_t) \, dt + \sigma(t, X_t) \, dB_t,
\]
where:
\begin{itemize}
  \item \( B_t \) is a standard Brownian motion;
  \item \( \mu(t, x) \) is the \textbf{drift coefficient}, governing the deterministic trend of the process;
  \item \( \sigma(t, x) \) is the \textbf{diffusion coefficient}, governing the stochastic fluctuations;
  \item The function \( \mu \) and \( \sigma \) are assumed to be sufficiently regular (e.g., measurable and satisfying growth and Lipschitz conditions).
\end{itemize}

\textbf{Examples.}
\begin{itemize}
  \item The Brownian motion itself: \( dX_t = dB_t \Rightarrow \mu = 0, \sigma = 1 \);
  \item The Ornstein–Uhlenbeck process: \( dX_t = -\theta X_t \, dt + \sigma \, dB_t \);
  \item The Geometric Brownian motion (used in finance): \( dX_t = \mu X_t \, dt + \sigma X_t \, dB_t \).
\end{itemize}

\vspace{1em}
\textbf{Why Classical Calculus Fails.}  
In deterministic analysis, we use classical tools such as differential calculus and chain rule (i.e., \( df(x(t)) = f'(x(t)) dx(t) \)).  
However, Brownian motion paths are almost surely nowhere differentiable and have infinite total variation over any interval. Therefore, classical calculus cannot be applied directly.

\vspace{1em}
\textbf{Why We Need Itô Calculus.}
\begin{itemize}
  \item \textbf{Irregularity of paths:} Since \( B_t \) has infinite variation, we must define a new type of integral: the \textbf{Itô integral}, which is well-defined for integrands adapted to the filtration generated by \( B_t \).
  \item \textbf{Modified chain rule:} The classical chain rule is replaced by \textbf{Itô’s formula}, which accounts for the stochastic nature of the process. For a twice differentiable function \( f(t, X_t) \):
  \[
  df(t, X_t) = \left( \frac{\partial f}{\partial t} + \mu \frac{\partial f}{\partial x} + \frac{1}{2} \sigma^2 \frac{\partial^2 f}{\partial x^2} \right) dt + \sigma \frac{\partial f}{\partial x} \, dB_t.
  \]
  \item \textbf{Foundation for stochastic models:} Itô calculus provides the mathematical foundation to define, solve, and analyze SDEs, which are essential in modeling random dynamics in physics, biology, economics, and especially in financial mathematics.
\end{itemize}

\textbf{Conclusion.}  
The study of stochastic processes—especially those driven by Brownian motion—requires a new calculus that respects their randomness and irregularity. Itô calculus fills this gap, making it possible to define integrals and derive dynamics of functions of stochastic processes.