\argomento{Stochastic Differential Equations (SDEs)}

\subsection{Definition.}  
A \textbf{Stochastic Differential Equation (SDE)} is an equation of the form
\[
dX_t = \mu(t, X_t) \, dt + \sigma(t, X_t) \, dB_t,
\]
where:
\begin{itemize}
    \item \( X_t \) is the unknown stochastic process,
    \item \( \mu(t, X_t) \) is the drift coefficient (deterministic or stochastic),
    \item \( \sigma(t, X_t) \) is the diffusion coefficient,
    \item \( B_t \) is a standard Brownian motion.
\end{itemize}

The solution to an SDE is defined as an Itô process that satisfies the integral version:
\[
X_t = X_0 + \int_0^t \mu(s, X_s) \, ds + \int_0^t \sigma(s, X_s) \, dB_s.
\]

\subsection{Examples of Financially Relevant SDEs}

\begin{itemize}
    \item \textbf{Geometric Brownian Motion (Black–Scholes Model):}
    \[
    dS_t = \mu S_t \, dt + \sigma S_t \, dB_t.
    \]
    It models the price of a risky asset under constant drift and volatility. The solution is:
    \[
    S_t = S_0 \exp\left( \left( \mu - \frac{1}{2} \sigma^2 \right)t + \sigma B_t \right).
    \]

    \item \textbf{Ornstein–Uhlenbeck Process (Mean-Reverting Process):}
    \[
    dX_t = \theta(\mu - X_t) \, dt + \sigma \, dB_t.
    \]
    This process models quantities that revert to a long-term mean \( \mu \) over time. Often used for interest rates or volatility.

    \item \textbf{Vasicek Interest Rate Model:}
    \[
    dr_t = a(b - r_t) \, dt + \sigma \, dB_t.
    \]
    This is an Ornstein–Uhlenbeck process applied to interest rates. It can become negative due to its Gaussian nature.

    \item \textbf{Cox–Ingersoll–Ross (CIR) Model:}
    \[
    dr_t = a(b - r_t) \, dt + \sigma \sqrt{r_t} \, dB_t.
    \]
    This is a mean-reverting model that ensures interest rates remain non-negative (as long as \( 2ab \geq \sigma^2 \)).

    \item \textbf{Heston Model for Stochastic Volatility:}
    \[
    \begin{cases}
    dS_t = \mu S_t \, dt + \sqrt{v_t} S_t \, dB_t^S, \\
    dv_t = \kappa(\theta - v_t) \, dt + \xi \sqrt{v_t} \, dB_t^v,
    \end{cases}
    \]
    with \( \text{Cov}(dB_t^S, dB_t^v) = \rho \, dt \). This two-dimensional SDE models asset prices with stochastic volatility.
\end{itemize}

SDEs are fundamental in financial modeling for capturing the randomness inherent in asset prices, interest rates, and volatility, and are the basis for option pricing, portfolio optimization, and risk management.